Using Monte Carlo Algorithms to Achieve Resiliency \& Performance at Scale for Linear and Nonlinear Solver Applications

### 2016-01-26 doc/ecp/pres.tex
Presentation update.

\documentclass{beamer}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage[compatibility=false]{caption}
\usepackage{tikz}
\usepackage{ragged2e}
\usepackage{algorithm,algorithmic}

@@

@@

  The MCREX project has 3 objectives for the development of Monte Carlo solver
  methods for linear systems:
  \begin{itemize}
  \item Analyze and improve the iterative performance of Monte Carlo
    algorithms for linear systems
  \item Analyze and characterize the resilient aspects of Monte Carlo
    algorithms for linear systems
  \item Develop and model highly scalable parallel algorithms for the
    implementation of Monte Carlo algorithms for linear systems
 \end{itemize}
 The end goal is to develop a set of resilient, scalable algorithms for the
 solution of linear systems.  We have made significant progress in each of
 these areas as summarized below.

@@

\begin{frame}{Summary}

  Much of the first 2 years of the project were attempting to resolve
  \textit{algorithmic} challenges related to solver performance and efficiency
  \begin{itemize}
  \item Iterative schemes to improve efficiency
  \item Analysis to determine convergence
  \item Development and application of effective preconditioners
  \end{itemize}

@@

  We have developed two open source software packages that contain
  production-quality versions of these solvers (MPI and GPU).

@@

\begin{frame}{Monte Carlo for Linear Systems}
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is the Richardson
      iteration matrix
      \vfill
    \item Build the Neumann series stochastically
  \end{itemize}

@@

  \begin{itemize}
  \item Define a sequence of state transitions
  \end{itemize}
  \vspace*{-0.1in}
  \[
  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
  \rightarrow i_{k}
  \]

@@

\begin{frame}{Limitations of Monte Carlo}
  \begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
    \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
    \begin{figure}
      \centering
      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
    \end{figure}
  \item There is a potentially restrictive limit on the spectral properties of
    $\mathbf{H}$
    \begin{equation*}
      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
      \frac{H_{ij}^2}{P_{ij}}
    \end{equation*}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Monte Carlo Synthetic Acceleration}
  \begin{itemize}
  \item Address issue of convergence constrained by CLT
    \vfill
  \item Can be abstracted as a general linear solver
    \vfill
  \item Combine with Richardson iteration as a ``smoother'' in between
    Monte Carlo steps:
    \begin{align*}
      \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
      \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
    \end{align*}
  \item We still must address issue of convergence constraint by spectral radius
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}{Polynomial Methods -- Summary}
  \begin{itemize}
    \item Using history length cutoff rather than weight cutoff can lead to
      large reductions in iteration counts
      \vfill
    \item A good approximation to a few terms in the Neumann series
      performs better than a statistically noisy approximation to
      full series
      \vfill
    \item Significant reduction in iteration counts are possible using
      alternate polynomials, but partially offset by increase in number
      of histories required
      \begin{itemize}
        \item May be very beneficial from resiliency and parallel efficiency
          standpoints
        \item Will be re-evaluated in future efforts
      \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Preconditioning}
  \begin{itemize}
    \item Preconditioning needed to address convergence constraints due to
      spectral radius conditions
    \item Preconditioning for Monte Carlo introduces nonstandard considerations
      \begin{itemize}
      \item Elements of preconditioned matrix must be explicitly available,
        sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity in
        $\mathbf{M^{-1}A}$
        \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
          may not follow ``standard'' trends
      \end{itemize}
  \end{itemize}
  \vfill

@@

\begin{frame}{Domain Decomposed Monte Carlo}

@@

\begin{frame}{Replication}

  Different batches of Monte Carlo samples can be combined in summation via
  superposition if they have different random number streams.  Consider each
  of these batches independent \textit{subsets} of a Monte Carlo operator
  where now the operator can be formed as a general additive decomposition of
  $N_S$ subsets:

  \[
  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
  \]

  We replicate the linear problem and form each subset on a different group of
  parallel processes. Applying the subsets to a vector requires an AllReduce
  to form the sum. Each subset is domain decomposed.
  \vfill
  This provides a mechanism to support \textit{resiliency}; the replicated
  component provides {\bf useful} work, not just redundancy

@@

\begin{frame}{Parallel Test - Simplified $P_N$ ($SP_N$) Assembly Problem}
\small
\begin{columns}

@@

\begin{frame}{Scaling - Domain Decomposition}

  \begin{figure}
    \centering
    \begin{tabular}{cc}
      \subfigure[Strong time]{\includegraphics[width=1.8in]{strong_dd_time}} &
      \subfigure[Strong efficiency]{\includegraphics[width=1.8in]{strong_dd_eff}} \\
      %%
      \subfigure[Weak time]{\includegraphics[width=1.8in]{weak_dd_time}} &
      \subfigure[Weak efficiency]{\includegraphics[width=1.8in]{weak_dd_eff}} \\
    \end{tabular}
  \end{figure}
\end{frame}

@@

\begin{frame}{Scaling - Domain Replication}

@@

  \begin{figure}
    \centering
    \begin{tabular}{cc}
      \subfigure[Strong time]{\includegraphics[width=2in]{strong_rep_time}} &
      \subfigure[Strong efficiency]{\includegraphics[width=2in]{strong_rep_eff}} \\
      %%
    \end{tabular}
  \end{figure}

@@

\begin{frame}{GPU Optimization}

@@

@@

\begin{frame}{Continued Research is Necessary}

  The following important questions need to be answered before MC methods can
  be applied to linear systems as a ``standard'' solver:

  \begin{itemize}
  \item Automatically determine history termination criteria
    \begin{itemize}
    \item We have investigated adaptive schemes based on variance
    \item However, variance is not directly correlated with error
    \end{itemize}
  \item Identify classes of matrices and preconditioners that satisfy
    convergence constraints
    \begin{itemize}
    \item This has been done for SDD and GDD
    \end{itemize}
  \item Demonstrate resiliency to hard MPI errors through replicated domain
    parallelism
  \item Further investigate non-linear, time-dependent problems
    \begin{itemize}
    \item This was part of the original proposal
    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
      dominant problems for which MCSA has been shown to be effective
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{MCSA Implementations}

  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
  \begin{itemize}
  \item Massively parallel (MPI) implementation of domain decomposed MCSA
    algorithm
  \item Implemented in Trilinos framework and callable through Trilinos'
    abstract numerical algorithms (Thyra, Stratimikos) interface
  \item Works with any Tpetra row matrix
  \end{itemize}
  \vfill
  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
  \begin{itemize}
  \item GPU-accelerated on-node MC solve
  \item Domain-local MC solves implemented for distributed systems as an
    additive Schwartz method
  \end{itemize}
  \vfill
  All software is BSD open source.
  \vfill
  Uses TriBITS (Trilinos) open source build system
  \vfill
  Both software packages contain a complete suite of unit-tests integrated
  through GoogleTest (Alea) or TeuchosTest (MCLS)
\end{frame}

@@

\begin{frame}{Connection to ECP Requirements}

@@

\begin{frame}{Inter-Operatibility Example}

  Using the MCLS interface to Trilinos, it is straightforward to implement
  MCSA in other applications
  \vfill
  We implemented MCSA in the production CFD code, Drekar (SNL) using a
  Newton-MCSA algorithm:
  \begin{algorithm}[H]
    \footnotesize
    \begin{algorithmic}[1]
      \STATE $k := 0$
      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
        ||\ve{F}(\ve{u}^{0})||$}
      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
      \COMMENT{Automatic differentiation - Panzer}
      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
      \COMMENT{Solve for the Newton correction with MCSA}
      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
      \STATE $k \leftarrow k+1$
      \ENDWHILE
    \end{algorithmic}
    \caption{FANM}
  \end{algorithm}
  \vfill
  Because MCLS was built with Trilinos abstract numerical algorithms this
  integration was straightforward and we can test MCLS in a whole new problem
  space
  \vfill
  We are also able to utilize all of Trilinos' preconditioners

@@

\begin{frame}{Thermal Convection Cavity Results}

@@

    \begin{column}{0.4\textwidth}

      \small{
        \begin{itemize}
        \item Weaker preconditioning at Ra=\sn{1}{5} case
          \medskip
        \item Constant forcing term for Ra=\sn{1}{6} case
        \end{itemize}
      }

      \bigskip

      \tiny{
        \begin{table}[h!]
          \begin{center}
            \begin{tabular}{lc}\hline\hline
              \multicolumn{1}{l}{Benchmark}&
              \multicolumn{1}{c}{NK Speedup}\\
              \hline
              Ra=\sn{1}{3} & 338 \\
              Ra=\sn{1}{4} & 336 \\
              Ra=\sn{1}{5} & 346 \\
              Ra=\sn{1}{6} & 465 \\
              \hline\hline
            \end{tabular}
          \end{center}
          \caption{Newton-Krylov speedup over FANM.}
        \end{table}
      }

    \end{column}
  \end{columns}

@@

  \underline{\bf XPRESS}
  \begin{itemize}
    \setlength{\itemsep}{-0.1\baselineskip}
  \item We are currently implementing a CUDA transport kernel with scheduling
    optimized by HPX as part of XPRESS
  \item We have already compared OpenMP versus HPX for on-node MC particle
    transport algorithms
    \begin{figure}
      \centering
      \begin{tabular}{cc}
        \subfigure[time]{\includegraphics[width=1.8in]{time}}
        &
        \subfigure[efficiency]{\includegraphics[width=1.8in]{tallies}}
      \end{tabular}
    \end{figure}
  \item All code implementations are available in the open source Profugus
    mini-app (\underline{github.com/ORNL-CEES/Profugus})
  \end{itemize}

@@

>>>>

\documentclass{beamer}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage[compatibility=false]{caption}
\usepackage{tikz}
\usepackage{ragged2e}
\usepackage{algorithm,algorithmic}

@@

% Phantom minus sign for helping with alignment of matrices
\newcommand{\phmin}{\ensuremath{\phantom{-}}}

@@

  The MCREX project has 3 objectives for the development of Monte Carlo solver
  methods for linear systems:
  \begin{itemize}
  \item Analyze and improve the iterative performance of Monte Carlo
    algorithms for linear systems
  \item Analyze and characterize the resilient aspects of Monte Carlo
    algorithms for linear systems
  \item Develop and model highly scalable parallel algorithms for the
    implementation of Monte Carlo algorithms for linear systems
 \end{itemize}
 The end goal is to develop a set of resilient, scalable algorithms for the
 solution of linear systems.

 \underline{Budget}: 300K/year (ORNL), 70K/year (Emory)

@@

\begin{frame}{Project Achievements}

  Much of the first 2 years of the project was spent resolving
  \textit{algorithmic} challenges related to solver performance and efficiency
  \begin{itemize}
  \item Iterative schemes to improve efficiency
  \item Analysis to determine convergence
  \item Development and application of effective preconditioners
  \end{itemize}

@@

  We have developed two open source software packages that contain
  parallel, production-quality versions of these solvers (MPI and GPU).

@@

\begin{frame}{Monte Carlo for Linear Systems}

  First reference, G.E. Forsythe and R.A. Leibler. Matrix Inverstion by a
  Monte Carlo Method. \textit{Math. Tables and Comp.}, {\bf 4}(31), 127--129,
  1950 (credited to von Neumann and Ulam, 1940s).
  \vfill
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is the Richardson
      iteration matrix
      \vfill
    \item Build the Neumann series stochastically
  \end{itemize}

@@

  \begin{itemize}
  \item Define a sequence of state transitions
  \end{itemize}
  \[
  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
  \rightarrow i_{k}
  \]

@@

\begin{frame}{Limitations of Monte Carlo---Efficiency}
  \begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
  \item Central limit theorem is barrier to accurate solutions
    \begin{figure}
      \centering
      \includegraphics[width=3in]{../NCState_seminar_2014/Ifiss_ConvDiff}
    \end{figure}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Limitations of Monte Carlo---Convergence}

  \begin{itemize}
  \item Because we are using MC to estimate terms in the Neumann series, the
    spectral radius of the iteration matrix must be
    \[
    \rho(\mathbf{H}) < 1
    \]

  \item Additionally, because we are estimating these terms stochastically,
    the relationship of the variance of the estimates to the error imposes an
    additional restriction 
    \[
    \rho(\hat{\mathbf{H}}) < 1\:, 
    \]
    where 
    \[
    \hat{H}_{ij} = \frac{H_{ij}^2}{P_{ij}}
    \]
    and the probabillities come from the decomposition of $\mathbf{H}$
    \[
    \mathbf{H} = \mathbf{P}\circ\mathbf{W}
    \]
  \end{itemize}
\end{frame}

@@

\begin{frame}{Can Monte Carlo Convergence be Improved?}

  \underline{Monte Carlo Synthetic Acceleration}
  \vfill
  \begin{itemize}
  \item Building on Halton's Sequential Monte Carlo, we obtain Monte Carlo as
    a corrective term as opposed to a direct solver to achieve exponential
    convergence
  \item Form iterative method in which Richardson iteration is a ``smoother''
    in between Monte Carlo steps:
    \begin{align*}
      \mathbf{x}^{k+1/2} &= (\ve{I} - \ve{A})\mathbf{x}^k + \ve{b} 
      && \text{Richardson Iteration}\\
      %% 
      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} && 
      \text{compute residual}\\
      %% 
      \hat{\mathbf{A}} \mathbf{\delta}_{MC} &= \mathbf{r}^{k+1/2}
      && \text{solve using MC}\\
      %%
      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta_{MC} 
      && \text{update} 
      %% 
    \end{align*}
  \item This scheme is still constrained by the spectral limits on $\ve{H}$
  \end{itemize}
  \vfill\centering
  \fbox{MCSA is a general linear solver}

@@

\begin{frame}{Why limit to the Neumann series polynomial?}
  \begin{itemize}
  \item Chebyshev or GMRES polynomials are viable alternatives
  \vfill
  \item A good approximation to a few terms in the polynomial
    performs better than a statistically noisy approximation to
    full series
  \vfill
  \item Using history length cutoff rather than weight cutoff can lead to
    large reductions in iteration counts
  \vfill
  \item Significant reduction in iteration counts are possible using
    alternate polynomials, but partially offset by increase in number
    of histories required
  \end{itemize}
  \vfill

  \begin{table}
    \caption{Polynomial MCSA - Shifted Laplacian}
    \centering
    \begin{tabular}{cccc}
      \toprule
      Polynomial & Polynomial & MCSA & Histories per \\
      Type & Order & Iterations & Iteration \\
      \midrule
      Neumann & 4 & 502 & 500 \\
      Chebyshev & 4 & 246 & 16000 \\
      GMRES & 4 & 260 & 2000 \\
      \bottomrule
    \end{tabular}
  \end{table}

@@

\begin{frame}{Enabling Convergence through Preconditioning}
  \begin{itemize}
    \item Preconditioning needed to address convergence constraints due to
      spectral radius conditions
    \item Preconditioning for Monte Carlo introduces nonstandard considerations
      \begin{itemize}
      \item Elements of preconditioned matrix must be explicitly available,
        sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity in
        $\mathbf{M^{-1}A}$
        \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
          may not follow ``standard'' trends
      \end{itemize}
  \end{itemize}
  \vfill

@@

\begin{frame}{Monte Carlo Scalability---Domain Decomposed Monte Carlo}

@@

\begin{frame}{Combining History Decomposition with Domain Decomposition}

  \begin{itemize}
  \item Different batches of Monte Carlo samples can be combined in summation
    via superposition if they have different random number streams
    \vfill
  \item The new operator can be formed as a general additive
    decomposition of $N_S$ subsets:
    \[
    \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
    \]
    \vfill
  \item We replicate the linear problem and form each subset on a different
    group of parallel processes.  Each subset is domain decomposed.  
    \vfill
  \item This provides a mechanism to support \textit{resiliency}; the
    replicated component provides {\bf useful} work, not just redundancy
  \end{itemize}
\end{frame}

@@

\begin{frame}{Application to Nuclear Reactor Analysis}
\small
\begin{columns}

@@

\begin{frame}{Scaling---Domain Decomposition}

  \begin{figure}
    \centering
    \begin{subfigure}{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{strong_dd_time}
      \caption{Strong scaling time}
    \end{subfigure}
    ~
    \begin{subfigure}{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{strong_dd_eff}
      \caption{Strong scaling efficiency}
    \end{subfigure}\\
    %%
    \begin{subfigure}{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{weak_dd_time}
      \caption{Weak scaling time}
    \end{subfigure}
    ~
    \begin{subfigure}{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{weak_dd_eff}
      \caption{Weak scaling efficiency}
    \end{subfigure}
  \end{figure}

@@

\begin{frame}{Scaling---History Replication + Domain Decomposition}

@@

  \begin{figure}
    \centering
    \begin{subfigure}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{strong_rep_time}
      \caption{Strong scaling time}
    \end{subfigure}
    ~
    \begin{subfigure}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{strong_rep_eff}
      \caption{Strong scaling efficiency}
    \end{subfigure}
  \end{figure}

@@

\begin{frame}{Monte Carlo GPU Algorithm for Linear Systems}

@@

\begin{frame}{Evolution of MC Linear Solver Software Products}

  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
  \begin{itemize}
  \item Massively parallel (MPI) implementation of domain decomposed MCSA
    algorithm
  \item Implemented in Trilinos framework and callable through Trilinos'
    abstract numerical algorithms (Thyra, Stratimikos) interface
  \item Works with any Tpetra row matrix
  \end{itemize}
  \vfill
  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
  \begin{itemize}
  \item GPU-accelerated on-node MC solve
  \item Domain-local MC solves implemented for distributed systems as an
    additive Schwartz method
  \end{itemize}
  \vfill
  All software is BSD open source.
  \vfill
  Uses TriBITS (Trilinos) open source build system
  \vfill
  Both software packages contain a complete suite of unit-tests integrated
  through GoogleTest (Alea) or TeuchosTest (MCLS)

@@

\begin{frame}{What research remains for DOE ECP integration?}

  The following important questions need to be answered before MC methods can
  be applied to linear systems as a ``standard'' solver in ECP applications

  \begin{itemize}
  \item Automatically determine history termination criteria
    \begin{itemize}
    \item We have investigated adaptive schemes based on variance
    \item However, variance is not directly correlated with error
    \end{itemize}
  \item Identify classes of matrices and preconditioners that satisfy
    convergence constraints
    \begin{itemize}
    \item This has been done for SDD and GDD
    \end{itemize}
  \item Demonstrate resiliency to hard MPI errors through replicated domain
    parallelism
  \item Identify optimal, tunable parameter values for performance in
    different applications
  \end{itemize}
\end{frame}

@@

\begin{frame}{Ties to Identified Requirements of ECP Applications}

@@

\begin{frame}{Inter-Operability---Connecting to Production Applications}

  MCLS implements Trilinos ANA interfaces making it straightforward to
  implement MCSA in other applications
  \begin{itemize}
  \item Exnihilo neutronics (shown before)
  \item Drekar (SNL) CFD
  \end{itemize}
  \vfill
  We implemented MCSA in Drekar using a Newton-MCSA algorithm via the Trilinos
  package NOX:
  \begin{algorithm}[H]
    \footnotesize
    \begin{algorithmic}[1]
      \STATE $k := 0$
      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
        ||\ve{F}(\ve{u}^{0})||$}
      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
      \COMMENT{Automatic differentiation - Trilinos FAD}
      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
      \COMMENT{Solve for the Newton correction with MCSA}
      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
      \STATE $k \leftarrow k+1$
      \ENDWHILE
    \end{algorithmic}
    \caption{FANM}
  \end{algorithm}
  \vfill
  \centering
  \fbox{Trilinos ANA enables application integration}

@@

\begin{frame}{Drekar-MCLS Thermal Convection Cavity Results}

@@

    \end{column}
    %%
    \begin{column}{0.4\textwidth}
      \vspace{-0.4in}
      \begin{figure}
        \centering
        \includegraphics[width=0.9\textwidth]{convection_isotherms}
        \caption{Steady state isotherms.}
      \end{figure}
    \end{column}

@@

  \underline{\bf XPRESS}
  \begin{itemize}
    \setlength{\itemsep}{-0.1\baselineskip}
  \item We are currently implementing a CUDA transport kernel with scheduling
    optimized by HPX as part of XPRESS
  \item We have already compared OpenMP versus HPX for on-node MC particle
    transport algorithms
    \begin{figure}
      \begin{subfigure}{0.4\textwidth}
        \centering
        \includegraphics[width=\textwidth]{time}
      \end{subfigure}
      ~
      \begin{subfigure}{0.4\textwidth}
        \centering
        \includegraphics[width=\textwidth]{tallies}
      \end{subfigure}
    \end{figure}
  \item All code implementations are available in the open source Profugus
    mini-app (\underline{github.com/ORNL-CEES/Profugus})
  \end{itemize}

@@

\begin{frame}{Forward Monte Carlo}
\begin{itemize}
  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
    \end{equation*}
  \item To compute solution component $\mathbf{x}_i$:
    \begin{itemize}
      \item Start a history in state $i$ (with initial weight of 1)
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_i$ based on current history weight
        and value of $\mathbf{b}_j$
    \end{itemize}
  \item A given random walk can only contribute to a single component of
    the solution vector
\end{itemize}
\end{frame}

\begin{frame}{Sampling Example (Forward Monte Carlo)}
  \begin{itemize}
    \item Suppose
  \begin{equation*}
    \mathbf{A} = \begin{bmatrix}
      \phmin 1.0 & -0.2 & -0.6 \\
      -0.4 & \phmin 1.0 & -0.4 \\
      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
       0.0 &  0.2 &  0.6 \\
       0.4 &  0.0 &  0.4 \\
       0.1 &  0.4 &  0.0 \end{bmatrix}
  \end{equation*}
    then
  \begin{equation*}
    \mathbf{P} = \begin{bmatrix}
       0.0 & 0.25 & 0.75 \\
       0.5 &  0.0 & 0.5 \\
       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
    \mathbf{W} = \begin{bmatrix}
       0.0 &  0.8 &  0.8 \\
       0.8 &  0.0 &  0.8 \\
       0.5 &  0.5 &  0.0 \end{bmatrix}
  \end{equation*}
    \vfill
    \item If a history is started in state $3$, there is a $20\%$ chance of
      it transitioning to state $1$ and an $80\%$ chance of moving to state
      $2$
  \end{itemize}
\end{frame}

\begin{frame}{Adjoint Monte Carlo}
\begin{itemize}
  \item Choose $\mathbf{P}$ and $\mathbf{W}$ such that
    $\mathbf{H}^{T} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ji}| }
      {\sum_{i=1}^{N} | \mathbf{H}_{ji} |}
    \end{equation*}
  \item To estimate solution:
    \begin{itemize}
      \item Start a history in random state $i$ by sampling from distribution
        determined by $\mathbf{b}$
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_j$ based on current history weight
    \end{itemize}
  \item A given random walk can contribute to many different components of
    the solution vector
\end{itemize}
\end{frame}
